Predicting Conflict Escalation

### 2023-05-03 docs/hes_final/article.qmd
Rewrite the results section.

We evaluated our models with both a research and a practical objectives in mind.

From a research perspective we were interested in feasibility of making accurate and generalizable predictions of
conflict escalation using publicly available text embeddings. We looked at the area under the ROC curve, F1 score, precision, recall and confusion matrix to design a holistic evaluation method at both fine-tuning/validation stage and final test set prediction stage. Despite conceptual differences in dataset creation and modeling techniques, both of the described above approaches produced comparative results.

@@

Further investigation showed that while overall numeric metrics look similar, the Transformer-based model generalized better across the globe. Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the
generalizability is a significant reason to prefer Transformer architecture and resampled dataset.

@@

While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making. With that in mind, cost of deployment is also an important factor for model evaluation. Average embedding dataset fits in memory while resampling article level embeddings requires additional data engineering
steps to be processed in batches. Both dataset size and corresponding model size make average embeddings based statistical
model more attractive from this perspective.

We think that both approaches produced promising results and the model choice depends on the developers' specific goals and available resources.

### Comparison with ...

>>>>

We evaluated the models against a set of common performance metrics (see Figures 4-9). Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results. Each of the models outperformed the other two
in a subset of these metrics. This suggests that the choice of the model architecture for this task would depend on specific application goals. We assume that UNICEF decision makers would
appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives. At the same time, if the false positive rate is too high the signal would become too
noisy for the tool to be useful. Given these assumptions and the class imbalance of the original
dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for
class imbalance) seem to be most relevant metrics. Two models using country features seem to perform
better on these two metrics.

In this work we defined the target in a way that was not used in prior literature, we cannot directly
compare our models' performance to others. The most similar work used a presense or absense
of any armed conflict related fatality as their binary target  [@10.1093/jeea/jvac025]. They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in peaceful countries) with their text based (LDA topic retrieval) model. Our best ROC_AUC score is 69% but the comparison is not very meaningful due to difference in targets.

@@

Further investigation showed that while overall numeric metrics look similar and there seems to be a slight performance advantage in using country data in addition to text features, the Transformer-based text-only model generalized better across the globe. Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the
generalizability is a significant reason to prefer Transformer architecture and resampled dataset.

@@

While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making. With that in mind, cost of deployment is also an important factor for model evaluation. 
Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings in two other proposed methods requires additional data engineering
steps to be processed in batches.

We think that all three approaches produced promising results and the model choice depends on the developers' goals, performance metrics and available resources.
